---
title: "Survival Analysis with R"
author: "Sean Walsh <sw23993@my.open.ac.uk>"
date: "15 January 2019"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE
  pdf_document: default
---

```{r setup, include = F}
# Set global options for code chunks and load required R packages and data
library(dplyr)
library(survival)
library(muhaz)
library(broom)
library(ggplot2)
data("lung") # Loads data but does not attach (I prefer not to attach in case variable names match across datasets)

knitr::opts_chunk$set(echo = T, # Show code chunks in R Markdown document
                      tidy  = F,
                      cache = F,
                      message = F,
                      warning = F)

# Rename levels in the status variable (necessary for functions used in this script)
lung$status <- if_else(lung$status == 2, 1, 0)
```


# Introduction

Survival analysis is a branch of statistics used for data analysis when the outcome of interest is time until an event occurs (i.e. **time-to-event data**). In bio-statistics, the event typically refers to death but it can be any outcome where the time until it occurs is of interest; it is the study of time between entry into observation and follow-up. 

Two-dimensional data such as these provide more information than binary data alone. For example, imagine a study is conducted which examines component failure. Component A has a failure rate of 22% after three years while Component B has a failure rate of 17% after three years. If all A components failed within one year but B components failed at various points across the three-year trial period, examining failure rate alone will not tell the full story. More insight is gained when including time until failure and this is especially true when dealing with treatment effects.

Survival analysis is widely applicable and can be used, for example, to study life expectancy, time until equipment failure, time until an earthquake or other environmental event, time until athlete retirement or time until stock market crash. The historical origin of survival analysis is slightly ambiguous but it is thought to have first emerged centuries ago. It was only after World War II that survival analysis, stimulated by an interest in the reliability of military equipment, became more widespread. A good introduction can be found at <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6110618/>.  


# Example Data

The clinical data set used for this workshop is the **NCCTG Lung Cancer Data**. The data set contains 228 observations and 10 variables. All ten variables are of numeric data type and the longest observation period is 2.8 years. Documentation on the data set can be obtained with `?lung`.

```{r lung data}
# Inspect the dataset structure 
str(lung)

# Longest observation time (in years)
max(lung$time) / 365
```

With survival analysis in mind, let's imagine that we are interested in the time until death in the clinical population of advanced lung cancer patients and we have obtained this sample of data. The event (death) is represented by the `status` variable and the observed survival time by the `time` variable. For a description of the encoding used within variables, we can refer to the help documentation at any time.


# The Survival Function

In survival analysis, the objective is to understand and potentially model the underlying distribution of survival times (i.e. the time until the event occurs). Two common ways of describing this distribution are the **survival function**, $S(t)$ and the **hazard function**, $h(t)$. 
The survival function gives the probability that a subject survives beyond some time or, in other words, the proportion of subjects who are event-free at some time. The survival function, $S(t)$, is a non-parametric estimator of survival at time $t$ and is defined as

$$
S(t_i) = S(t_{i-1})(1-\frac{n_i}{R_i})
$$

Note that $S(t_{0}) = 1$ always and survival probability decreases **monotonically** with time $t$. Given its entirely non-increasing nature, $S(t)$ is a monotonic function.


# The Hazard Function

The hazard function $h(t)$ represents the instantaneous rate of failure at any time conditional on survival up until that time. It can be defined as the survival time density $f(t)$ at time $t$ over the survival probability at time $t$. 

The hazard and survival functions provide alternative but equivalent descriptions of the distribution of survival times in a sample. If we obtain the survival function, we can simply differentiate to obtain the probability density function and then calculate the corresponding hazard function using the relationship described earlier.

The **Kaplan-Meier** approach estimates the survival function using `survfit()` while `muhaz()` gives us smoothed estimates of the hazard function. The `pehaz()` functions can be used to obtain histogram estimates of the hazards. The histograms use a piece-wise 
exponential hazard function which estimates the hazard in each bin as the number of events in that bin divided by the total follow-up time in that bin. 


# Survival Time Distribution

If some of you attended an R Generation data visualisation workshop, you will be familiar with some of the following code 
and plot types. This first plot is a scatter plot of the number of events (deaths) against follow-up time observed in the study. 

```{r survival times}
# Create a Kaplan-Meier fit on-the-fly and plot the survival time distribution
survfit(Surv(time, status) ~ 1, 
        data = lung) %>%
  tidy() %>%
  filter(n.event > 0) %>%
  ggplot(aes(x = time,
             y = n.event)) +
  geom_jitter(alpha = 0.75,
              colour = "red") +
  ylab("Events") +
  xlab("Time (days)") +
  ggtitle("Distribution of survival times in the NCCTG data set",
          subtitle = "Source: survival R package") +
  theme(text = element_text(face = "bold",
                            family = "AvantGarde"))
```

A histogram gives us a better view of the survival time distribution. Here is a base R histogram which shows that shorter survival times are more probable, particularly in the first 400 days.

```{r hist}
# Histogram of survival times
lung %>%
  filter(status == 1) %>% # Subset only subjects that experienced the event
  pull(time) %>% # Pull the time variable as a vector
  hist(main = "Survival time distribution in NCCTG data",
       xlab = "Time (days)",
       col = "red")
```

Next, a probability density function (PDF) plot is created. We know that shorter survival times are more probable in this data set and this comes through in the PDF plot as well. We will obtain a median survival time in a later section to see how it relates to the histogram and probability density function.

```{r pdf}
# The probability density function
lung %>%
  filter(status == 1) %>% 
  pull(time) %>%
  density() %>%
  plot(main = "PDF of survival times in NCCTG data",
       xlab = "Time (days)")
```

The histogram estimate of the hazard function obtained using `pehaz()` is shown below. 

```{r hazard function}
# The baseline hazard function estimate
plot(pehaz(lung$time,
           lung$status,
           max.time = max(lung$time)),
     xlab = "Time (days)",
     main = "Hazard function estimate based on NCCTG data",
     col = "red")
```


# The Kaplan-Meier Estimator

If we create a null Kaplan-Meier survival function for these data, we can inspect a general median survival time and survival curve. Let us now estimate the survival function using no covariates

```{r Kaplan-Meier}
# Create a Kaplan-Meier survival function fit
km_fit <- survfit(Surv(time, status) ~ 1,
                  data = lung)
```

and then plot the resulting object

```{r Kaplan-Meier plot}
# Plot the Kaplan-Meier estimator
plot(km_fit,
     xlab = "Time (days)",
     ylab = "Survival probability",
     main = "Kaplan-Meier estimator based on NCCTG data")
```

The result is a **monotonic** function where survival probability is entirely non-increasing over time. The 95% confidence interval at each time point is demarcated by the dashed lines. The **median survival time**, which is the point by which 50% of subjects have experienced the event, can be obtained by simply calling the survival fit object. 

```{r median survival time}
# Median survival time
km_fit
```

The median survival time based on these data is 310 days. Using the 95% confidence interval, a reasonable estimate of survival time would be 285 to 363 days. Note that this is an estimate based on the null Kaplan-Meier fit (i.e. no covariates)..


# Survival Across Groups

We plotted the null Kaplan-Meier fit earlier but what if we want to explore survival across groups? We can easily do this using the formula method. Let's compare survival across genders by introducing the `sex` variable as a single covariate.

```{r group comparison}
# Kaplan-Meier survival estimators by gender
km_fit_gender <- survfit(Surv(time, status) ~ sex,
                         data = lung)

# Plot the survival functions
plot(km_fit_gender,
     main = "Kaplan-Meier estimators by gender\n based on NCCTG data",
     ylab = "Survival probability",
     xlab = "Time (days)",
     lty = c(1, 3),
     col = c("orange", "blue"))

# Add a legend
legend(cex = .65,
       legend = c("Male", "Female"),
       lty = c(1,3),
       col = c("orange", "blue"),
       bty = "o",
       x = 50,
       y = 0.2)
```

The survival curves are clearly different with females showing better survival over time. The survival curve for females consistently runs above that for males across the observational period. The median survival time is 270 days for males and 426 days for females. Upper and lower 95% confidence interval levels are also present.

```{r group median survival times}
# Summary of fit
km_fit_gender
```

The log-rank test is the statistical test of choice when comparing survival across groups. The observed difference in survival across genders is statistically significant (*p* < 0.001).

```{r log-rank test}
# Perform the log-rank statistical test 
survdiff(Surv(time, status) ~ sex,
         data = lung)
```


# Cox Proportional Hazards Regression

The Kaplan-Meier approach is excellent when comparing survival across groups. However, if we want to investigate the effect of multiple covariates simultaneously then a regression strategy is required. The **Cox Proportional Hazards** regression model is a popular technique which allows the inclusion of both continuous and categorical data as covariates. Its major assumption is proportional hazards and it models the hazard function discussed earlier. The proportional hazards assumption should be thoroughly checked using Kaplan-Meier estimators or the `cox.zph()` function. The survival function can be constructed using the Cox Proportional Hazards model output. 

A notable strength of the model is that it assumes no particular distribution for the baseline hazard. Accelerated Failure Time (AFT) models are another good approach when survival modelling but they require a parametric form specified for the baseline hazard function.

In the Cox model, a baseline hazard $h_0$ is specified and the model is a linear combination of covariates. The effect of the covariates is to multiplicatively shift the baseline hazard and these shifts are constant over time under the proportional hazards assumption

$$
h(t_i)= h(t_{_0})exp(\beta_1x_1 + ... + \beta_px_p)
$$

The modelled hazard can be greater or less than the arbitrarily defined baseline. A hazard **> 1** means the hazard is greater than the baseline and the survival curve will run below the baseline curve (i.e. worse survival). A hazard **< 1** means the hazard is less than the baseline and the survival curve will run above the baseline curve (i.e better survival). A hazard of **0** means that the hazard is no different to the baseline hazard and the survival curve will run along the same path as the baseline curve. 

A few final points worth noting are that the Cox Proportional Hazards Regression model is a **relative** model and the **hazard ratios** are of utmost importance when interpreting the model output. Let's build a Cox Proportional Hazards regression model using the `coxph()` function with `sex` and `age` as covariates.

```{r cox ph regression}
# Make sex a factor (makes model interpretation easier)
lung$sex <- factor(lung$sex)

# Build a Cox PH regression model using sex and age as covariates
cox_ph_mdl <- coxph(Surv(time, status) ~ sex + age, # Additional covariates can be added as necessary
                    data = lung) 

# Retrieve the model output
summary(cox_ph_mdl)
```


# Model Interpretation

The Concordance, or $c$-index, is the proportion of pairwise comparisons correctly identified by the model and can be used as a measure of model strength. The $c$-index is similar to the AUC model validation measure used in machine learning applications. Good Cox Proportional Hazards regression models will have a $c$-index between 0.6 and 0.7. The baseline model, i.e. the random guess model, is represented by a $c$-index of 0.5.

The **hazard ratio** (exp(coef)) is the statistic of interest if its corresponding *p*-value is < 0.05. This important statistic compares the hazards of treatment and reference groups to quantify relative risk. The three global significance tests are asymptotically similar but the likelihood ratio test result is typically favoured. 

Finally, we can test the proportional hazards assumption using `cox.zph()`. If the *p*-value is > 0.05, then the proportional hazards assumption holds true. If proportional hazards is not present, then **stratifying** on that covariate may provide a solution.

```{r ph test}
# Test the proportional hazards assumption
cox.zph(cox_ph_mdl)
```


